Spatial Reasoning and Statistical Graphics

### 2014-10-20 Paper/Draft1/TestingVisualAptitude.Rnw
Moving demographic information and code around.

<<VisReasoningAnalysis,echo=FALSE, include=FALSE, fig.width=8,fig.height=8, out.width='0.9\\textwidth', message=F, warning=F>>=
lineup.summary$variable <- factor(lineup.summary$variable, labels=c("Card Rotation", "Figure Classification", "Paper Folding", "Visual Search"))

qplot(data=lineup.summary, x=value, y=lineup, geom="point", shape=I(1), size=I(3)) + facet_wrap(~variable, scales="free_x") + xlab("Scaled Score") + ylab("Scaled Lineup Score") + geom_smooth(method="lm") + theme_bw()
# same aspect ratio
ggsave(paste(imgdir, "fig-VisReasoningAnalysisCts.png", sep=""), width=6, height=6)

lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="verbal_skills", value=1))
lineup.summary.categorical <- rbind(lineup.summary.categorical, data.frame(id=1, lineup=NA, variable="math_skills", value=1))
lineup.summary.categorical$variable <- factor(lineup.summary.categorical$variable, labels=c("Age", "Sex", "Math/Science Research", "Statistics Class", "Calculus 1", "Verbal Skills", "Math Skills", "Art Skills", "AutoCAD Experience", "Video Game Hrs/Wk"))
lineup.summary.categorical$value[lineup.summary.categorical$value=="f"] <- "Female"
lineup.summary.categorical$value[lineup.summary.categorical$value=="m"] <- "Male"
lineup.summary.categorical$value[lineup.summary.categorical$value=="y"] <- "TRUE"
lineup.summary.categorical$value[lineup.summary.categorical$value=="n"] <- "FALSE"

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot") + facet_wrap(~variable, scales="free_x") + xlab("") + ylab("Scaled Lineup Score") + theme_bw() + geom_point(aes(x=value, y=lineup), shape=1, size=3)
ggsave(paste(imgdir, "fig-VisReasoningAnalysisCat.png", sep=""), width=6, height=6)
@

\begin{figure}[htbp]\centering
\includegraphics[width=0.75\textwidth]{fig-VisReasoningAnalysisCts}
\caption[Visual Aptitude Study Results]{Scatterplots of all test scores compared to participants' scores in the lineup tests. There is a relatively strong positive correlation between lineup score and scores on visuospatial reasoning tests.}\label{fig:visualaptitudects}
\end{figure}

@@

\begin{figure}[htbp]\centering
\includegraphics[width=\textwidth]{fig-VisReasoningAnalysisCat}
\caption[Visual Aptitude Study Results]{Sample demographic characteristics compared with lineup score. Sex (male), Calculus completion, and hours spent playing video games per week are all associated with a difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure}

There is no significant difference in lineup performance for participants of different age, self-assessed skill rating, previous participation in math or science research, completion of a statistics class, or experience with AutoCAD.
There is a significant difference between male and female performance on lineups; this is not particularly surprising, since men perform better on many spatial tests \citep{voyer1995magnitude} and performance on spatial tests is correlated with phase of the menstrual cycle in women \citep{hausmann2000sex}.
Completion of Calculus I is associated with increased performance on lineups. This may be related to general math education level, or it may be that success in both lineups and calculus requires certain visual skills. \newtext{This result is consistent with \citep{shah1995conceptual}, who found an association between  mathematical ability and performance on simple graph description tasks.}  There is also a significant association between hours of video games played per week and score on lineups, however, this association is not monotonic and the groups do not have equal sample size, so the conclusion may be suspect.

<<ttests.categ, echo=F, results='asis'>>=
tmp <- dlply(lineup.summary.categorical, .(variable), function(df){
    tmp <- anova(lm(lineup~factor(value, ordered=F), data=df))
    coefs <- as.data.frame(tmp)[1,]
    coefs$level <- gsub("factor(value, ordered = F)", unique(df$variable), rownames(coefs), fixed=T)
    rownames(coefs) <- NULL
    coefs <- coefs[,c(6, 1:5)]
    names(coefs) <- c("Variable", "DF", "Sum.of.Squares", "Mean.Squared", "F.value", "p.value")
    coefs
})

tmp <- rbind.fill(tmp)
tmp <- tmp[order(tmp$p.value, decreasing=F),]
print(xtable(tmp, digits=c(1, 1, 0, 3, 3, 2, 3), caption="Model results of each demographic variable compared with lineup score. Multiple testing issues aside, it appears that very few demographic variables (if any) are significantly associated with score on lineups among Iowa State undergraduate students."), include.rownames=F)
@

>>>>

<<VisReasoningCts,echo=FALSE, include=FALSE, eval=T, fig.width=8,fig.height=8, message=F, warning=F>>=
lineup.summary$variable <- factor(lineup.summary$variable, labels=c("Card Rotation", "Figure Classification", "Paper Folding", "Visual Search"))

qplot(data=lineup.summary, x=value, y=lineup, geom="point", shape=I(1), size=I(3)) + facet_wrap(~variable, scales="free_x") + xlab("Scaled Score") + ylab("Scaled Lineup Score") + geom_smooth(method="lm") + theme_bw()
@

\begin{figure}[htbp]\centering
\includegraphics[width=0.75\textwidth]{fig-VisReasoningCts-1}
\caption[Visual Aptitude Study Results]{Scatterplots of all test scores compared to participants' scores in the lineup tests. There is a relatively strong positive correlation between lineup score and scores on visuospatial reasoning tests.}\label{fig:visualaptitudects}
\end{figure}

@@
